### 2024-05-16 thesis/input_files/probabilistic.tex
May 16th 2024

We saw in the previous chapter how point forecasting relates to predicting the expected value of the target variable.
Conversely, in probabilistic forecasting we are interested in modelling a probability distribution over the target variable.
This approach enables us to account for prediction uncertainty and to make more informed decision.
\section{Quantile regression}
\input{input_files/qr.tex}

\section{Quantile forest}
Meinshausen \cite{meinshausen2006quantile} extends the idea of random forest \cite{breiman2001random} generalising it, the result is the quantile forest algorithm. Quantile forest allows us to estimate conditional quantiles in a non parametric fashion.
\\
In order to understand this algorithm, it is first necessary to cover the theory of decision trees and random forests.
\subsection{Decision trees}
Decision trees methods partition recursively the feature space in a set of binary rectangles and then fit a simple model in each of those partitions (the most straightforward is fitting just a constant).
To get started, we first split the space into two disjoint regions, then we model the response variable by the mean of the observed predicted variables with associated features falling in that specific region. Our goal is selecting the best features and best split point such that to achieve the best generalizing fit.
For a visualisation consider the pictures \ref{fig:elements_statistical_learning1} and \ref{fig:elements_statistical_learning2} taken from \cite{hastie2009elements}, where the decision tree algorithm is visualised for a regression problem with two independent variables $X_2$ and $X_1$.
\begin{figure}
    \includegraphics[width=0.5\textwidth]{images/elsii1.png}
    \caption{two-dimensional feature space partitioned by recursive binary splitting \cite{hastie2009elements}}
    \label{fig:elements_statistical_learning1}
  \end{figure}
\\
\begin{figure}
\includegraphics[width=1\textwidth]{images/elsii2.png}
\caption{partition tree and regression model \cite{hastie2009elements}}
\label{fig:elements_statistical_learning2}
\end{figure}
\\
Suppose now to partition the feature space into M regions $R_1,\dots,R_M$, then the model reads as follows.
\begin{equation}
    f(x)=\sum\limits_{m=1}^{M}c_m\mathbb{I}_{\{x \in R_m\}}
\end{equation}
\\
It follows that the function minimising the sum of squares is the one with $\hat{c}_m=mean(y_i|x_i \in R_m)$.
Finding the best split in terms of minimum sum of squares is computationally infeasible in practice. Therefore we approximate a solution by approaching the problem in a greedy fashion.
Let $j$ denote the splitting variable and $s$ be the the split point we define the two half planes
\begin{equation}
    R_1(j,s)=\{X|X_j\leq s\} \quad R_2(j,s)=\{X|X_j >s\}
\end{equation}
Then we search for the $s$ and $j$ that solve
\begin{equation}
    \min_{j,s}\left[\min_{c_1} \sum\limits_{x_i \in R_1(j,s)}(y_i-c_1)^2+\min_{c_2} \sum\limits_{x_i \in R_2(j,s)}(y_i-c_2)^2\right]
\end{equation}
The inner problem is easy, as already pointed out, we will have 
\begin{equation}
    \hat{c}_1=mean(y_i|x_i \in R_1) \quad \hat{c}_2=mean(y_i|x_i \in R_2)
\end{equation}
For the outer problem, we scan through all the (j,s) tuples and pick the best pair. 
Next, one or both of these regions from the previous step are split into two more regions; we recurse this process until some stopping condition is triggered (max number of branches, max deth of tree, threshold on the MSE, minimum number of observation in each leaf node).
Notice, this being a greedy algorithm implies that our final solution is guaranteed to be just a local optimum not a global one.
Even though their simplicity, these models have proved themselves to be really powerful. See figure \ref{fig:decision_tree} for a example where decision tree with different hyperparameters are fitted to a sine wave plus noise. The most popular among these model is the CART \cite{breiman2017classification} tree, its name comes from the fact that it can handle both classification and regression problems.
\begin{figure}
    \includegraphics[width=\textwidth]{images/decision_tree.png}
    \caption{decision tree regression}
    \label{fig:decision_tree}
    % image generated on coolab to be quick
\end{figure}

@@

% \section{Kernel methods}
\section{Kernel quantile regression}
The idea of quantile regression has been extended to kernel methods by Takeuchi et al. \cite{takeuchi2006nonparametric}.
There, they minimize a risk functional plus regularizer defined as follows.
\begin{equation}\label{eq:kqr_min1}
    R[f]:=\frac{1}{m}\sum\limits_{i=1}^{m}\rho_q(y_i-f(x_i))+\frac{\lambda}{2}\|g\|_\mathcal{H}^2
\end{equation}
where $f=g+b$, $g \in \mathcal{H}$ and $b \in \mathbb{R}$.
Using the link between RKHS and feature spaces, we can rewrite $f(x)=\langle w, \phi(x) \rangle_{\mathcal{H}}+b$. Moreover, note that the RKHS norm is defined as follows $\|f\|_{\mathcal{H}}=\inf\{\|w\|_{\mathcal{F}}:w\in \mathcal{F},f(x)=\langle w,\varphi (x)\rangle _{\mathcal{F}},\forall x\in X\}$, where $\mathcal{F}$ is the feature space.
 Doing so we obtain a minimization problem equivalent to minimizing equation \ref{eq:kqr_min1}.
\begin{equation}\label{eq:kqr_min2}
    \begin{aligned}
    \min_{w,b} \quad & C \sum \limits_{i=1}^{m}
    q(y_i-\langle w, \phi(x) \rangle_{\mathcal{H}}-b)+ (1-q)(-y_i+\langle w, \phi(x) \rangle_{\mathcal{H}}+b)+ \frac{1}{2}\|w\|_{\mathcal{H}}^2\\
    \end{aligned}
    \end{equation}
Note the division by $\lambda$ so that $C=\frac{1}{\lambda m}$.
\\
We can next rephrase the optimisation in \ref{eq:kqr_min2} by introducing the slack variables $\xi_i$ and $\xi_i^*$.
\begin{equation}\label{eq:kqr_min3}
    \begin{aligned}
        \min_{w,b,\xi_i,\xi_i^*} \quad & C \sum \limits_{i=1}^{m}
        q \xi_i+ (1-q)\xi_i^*+ \frac{1}{2}\|w\|_{\mathcal{H}}^2\\
    \textrm{s.t.} \quad & y_i-\langle w, \phi(x) \rangle_{\mathcal{H}}-b \leq \xi_i\\
    & -y_i+\langle w, \phi(x) \rangle_{\mathcal{H}}+b \leq \xi_i^*\\
      &\xi_i\geq0    \\
      &\xi_i^*\geq0    \\
    \end{aligned}
    \end{equation}
In order to make it more compact, we rewrite equation \ref{eq:kqr_min3} in matrix notation.
\begin{equation}\label{eq:kqr_min4}
    \begin{aligned}
        \min_{w,b\xi,\xi^*} \quad & C q \xi^\intercal \mathbb{1}+ C q (\xi^*)^\intercal \mathbb{1}+ \frac{1}{2}w^\intercal w\\
    \textrm{s.t.} \quad & y-\Phi^\intercal w -b \preceq \xi\\
    & -y+\Phi^\intercal w +b \preceq \xi^*\\
      &\xi\succeq0    \\
      &\xi^*\succeq0    \\
    \end{aligned}
    \end{equation}
Consider now, the lagrangian $\mathcal{L}$ associated to \ref{eq:kqr_min4}
\begin{equation}\label{eq:kqr_min5}
    \begin{aligned}
    L(w,b,\xi,\xi^*)= & C q \xi^\intercal \mathbb{1}+ C q (\xi^*)^\intercal \mathbb{1}+ \frac{1}{2}w^\intercal w- \alpha^\intercal(\xi - y+\Phi^\intercal w +b)
    - (\alpha^*)^\intercal(\xi^* +y-\Phi^\intercal w -b)
    \\
    & -\nu^\intercal \xi - (\nu^*)^\intercal \xi^*
\end{aligned}
\end{equation}
The next step is deriving its dual formulation, since it is easier and more efficient to solve. This because the dual problem has the useful property of being always convex.

@@

    The dual function associated to the lagrangian $\mathcal{L}(x,\lambda)$ is given by $g(v)=\inf_x \mathcal{L}(x,\lambda)$

@@

where $\lambda$ is called the lagrange multiplier of the optimization problem. Such dual formulation has an useful property, that is \begin{equation}\label{weak_duality}
    g(\lambda)\leq p^*

@@

where $p^*$ is the optimal value of your optimization problem.
Consider a simple lagrangian $\mathcal{L}(x,\lambda)=f(x)+\sum\limits_{i=1}^n \lambda_i r_i(x) +\sum\limits_{i=1}^{n} v_i h_i(x)$, where $r_i(x)$ are inequality constraints while $h_i(x)$ are equality constraints of the problem. Then it can be noted that, the lower bound on $p^*$ is non trivial only when the lagrange multiplier $\lambda \succeq 0$. Therefore, the idea is that by maximizing the dual function subject to the contraint $\lambda \succeq 0$, we can obtain an approximate or perfect solution to the primal problem.

@@

@@

Where relint $D$ is the relative interior of $ D:=\cap _{i=0}^{m}\operatorname {dom} (r_{i})$

@@

In our case we don't have any equality constraint, so we do not have to worry about the $h_i(x)=0$ term in \ref{slater_condition}. All we have to check is the convexity of our problem and that there exist a $x$ such that $r_i(x)<0$.

@@

This fact is easily checked by the fact that kernel matrices are by definition positive semidefinite.

@@

Next we check that Slater's condition holds. Considering first the two non negative constraints on $\xi$ and $\xi^*$, we conclude that $\xi$  and $\xi^*$ have to be greater or equal to zero for the existence of an $x$ satisfying Slater's condition. Thus, let us suppose that $0 \leq \xi \leq \alpha$ and $0 \leq \xi^* \leq \alpha$.

@@

        w=& \Phi^\intercal(\Phi \Phi^\intercal)^{-1} y\\
        b<& \alpha

@@

We then have

@@

        -\xi + y -\Phi\Phi^\intercal(\Phi \Phi^\intercal)^{-1}y-b<&0

@@

        -\xi^* - y +\Phi\Phi^\intercal(\Phi \Phi^\intercal)^{-1}y+b<&0

@@

We end this section with the derivation of the dual problem; that is the convex problem, we will solve in order to get the quantiles prdeiction of our quantile kernel algorithm.

@@

        g(\alpha, \alpha^*, \nu, \nu^*)=& \inf_x \mathcal{L}(x,\lambda)\\
    = & \inf_{\xi, \xi^*, w, b} \mathcal{L}(w,b,\xi,\xi^*)

@@

        \frac{\partial L}{\partial w}=0 \implies w=\Phi^\intercal(\alpha-\alpha^*)

@@

        \frac{\partial L}{\partial b}=0 \implies (\alpha-\alpha^*)^\intercal\mathbb{1}=0

@@

        \frac{\partial L}{\partial \xi}=0 \implies Cq \mathbb{1}-\alpha- \nu=0

@@

        \frac{\partial L}{\partial \xi^*}=0 \implies C(1-q)\mathbb{1} -\alpha^* -\nu^*=0

@@

As pointed out previously, the lower bound resulting from the dual formulation is non trivial only when the lagrange multipliers are $\succeq 0$. Looking at the last two equations of the system \ref{eq:lagrange_derivatives}, this implies the following two constraints $\alpha \in [0, Cq\mathbb{1}]$ and $\alpha^*\in [0, C(1-q)\mathbb{1}]$.

@@

@@

        g(\alpha, \alpha^*)=& \xi^\intercal(Cq\mathbb{1}-\alpha -\nu)+(\xi^*)^\intercal(C(1-q)\mathbb{1}-\alpha^*-\nu^*)-(\alpha-\alpha^*)^\intercal \Phi\Phi^\intercal(\alpha-\alpha^*)

@@

        & +(\alpha-\alpha^*)^\intercal y-(\alpha-\alpha^*)^\intercal b+\frac{1}{2}(\alpha-\alpha^*)^\intercal \Phi\Phi^\intercal(\alpha-\alpha^*)

@@

        g(\alpha, \alpha^*)=& 0+0-\frac{1}{2}(\alpha-\alpha^*)^\intercal \Phi\Phi^\intercal(\alpha-\alpha^*)+(\alpha-\alpha^*)^\intercal y-0

@@

        g(\alpha, \alpha^*)=& -\frac{1}{2}(\alpha-\alpha^*)^\intercal \Phi\Phi^\intercal(\alpha-\alpha^*)+(\alpha-\alpha^*)^\intercal y

@@

Defining $\textrm{a}=(\alpha-\alpha^*)$ and letting $K$ the kernel matrix, we have that the dual optimisation problem reads as follows

@@

        \max_{\textrm{a}} \quad & -\frac{1}{2}\textrm{a}^\intercal K\textrm{a}+\textrm{a}^\intercal y\\

@@

    C(q-1)\mathbb{1}\preceq \textrm{a}\preceq Cq\mathbb{1}\\
    &\textrm{a}^\intercal\mathbb{1}=1

@@

        \min_{\textrm{a}} \quad & +\frac{1}{2}\textrm{a}^\intercal K\textrm{a}-\textrm{a}^\intercal y\\

@@

    C(q-1)\mathbb{1}\preceq \textrm{a} \preceq Cq\mathbb{1}\\
    &\textrm{a}^\intercal\mathbb{1}=1

@@

    f(x)=\sum\limits_{i=1}^{n} \textrm{a}_i k(x_i, x)+b

@@

Since our optimisation problem possesses strong duality and it is differentiable in both the objective and the constraint, we have that it must satisfy the Karush Kuhn Tucker conditions, see section 5.5.3 \cite{boyd2004convex}.
Thanks to the KKT conditions on the primal optimisation problem we have that $f(x_i)=y_i \ \mathrm{for} \ \textrm{a}_i \not \in \{C(q-1), Cq\}$. 

@@

@@

    \lambda^* r_i(x^*)=&0, \quad i=1,\dots,n

@@

    \nabla L(x^*)=&0

@@

        \alpha_i(\xi_i-y_i+r_i)=&0

@@

        \alpha_i^*(\xi_i^*+y_i-r_i)=&0

@@

        \nabla L =&0

@@

        (Cq-\alpha_i) \xi_i=&0

@@

        (C(1-q)-\alpha_i) \xi_i^*=&0

@@

\begin{equation}
\begin{cases}
    \alpha_i=Cq, \ \alpha_i^*=0 & \implies \alpha_i-\alpha_i^*=Cq, \xi_i\leq 0, \xi^*=0\implies \xi_i-y_i+f_i=0 \\
    \alpha_i=0, \ \alpha_i^*=C(1-q) & \implies \alpha_i-\alpha_i^*=C(q-1), \xi_i= 0, \xi^*\leq 0 \implies \xi_i^*+y_i-f_i=0 \\
    0\leq \alpha_i\leq Cq, \ 0\leq \alpha_i^*\leq C(1-q) & \implies \xi_i=0,  \xi^*_i= 0 \implies -y_i+f_i=0, \ y_i-f_i=0 \\
\end{cases}
\end{equation}

Therefore in order to retrieve b we simply have to choose an index $i$ such that  $\textrm{a}_i \not \in \{C(q-1), Cq\}$
and let
\begin{equation}
    \begin{aligned}
    b=&y_i-\sum\limits_{i=1}^n \textrm{a}_i k(x_i,x)
    \end{aligned}
\end{equation}

@@

% tables
\begin{table}[!h]
\caption{Pinball loss Melbourne data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    Pinball loss & 11.278895 & 10.317612 & 10.340842 & \textbf{10.031708} \\
    \bottomrule
    \end{tabular}
\end{table}

\begin{table}[!h]
    \caption{Pinball loss quantile-wise Melbourne data}
    \begin{tabular}{lllll}
    \toprule
    Quantiles & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    0.100000 & 0.710644 & 0.549232 & 0.562888 & \textbf{0.540235} \\
    0.200000 & 1.155014 & 0.938561 & 0.946712 & \textbf{0.903213} \\
    0.300000 & 1.417805 & 1.212671 & 1.222407 & \textbf{1.173803} \\
    0.400000 & 1.540108 & 1.399925 & 1.409293 & \textbf{1.367943} \\
    0.500000 & 1.574957 & 1.517281 & 1.484589 & \textbf{1.455849} \\
    0.600000 & 1.525114 & 1.498608 & 1.495474 & 1.\textbf{447470} \\
    0.700000 & 1.397918 & 1.372183 & 1.362173 & \textbf{1.331503} \\
    0.800000 & 1.170140 & 1.115077 & 1.123195 & \textbf{1.096606} \\
    0.900000 & 0.787195 & \textbf{0.714075} & 0.734112 & 0.715085 \\
    \bottomrule
    \end{tabular}
\end{table}
As already pointed out, the quantile regression with q=0.5 corresponds to the standard regression problem, hence we can compare the proposed methods also in terms of the mean absolute error.
\begin{table}[!h]
\caption{Mean absolute error Melbourne data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    MAE & 3.253882 & 3.134805 & 3.095041 & \textbf{3.024336} \\
    \bottomrule
    \end{tabular}
\end{table}  
From these tables, we can see that kernel quantile regression outperformed the simple quantile regressor as well as the more complex models like quantile forest and gradient boosting quantile regression for the Melbourne temperatures dataset. Not only kernel quantile regression was the best in terms of total pinball loss but also the best in terms of each quantile pinball loss and mean absolute error.
Comparison has been carried out on further datasets, yielding similar conclusion as the one of above, to know more see \ref{appendix:quantile_regressor_extensive_comparison}.

We conclude this chapter with reporting the same kind of tables \ref{tab:kernel pinball comparison}, \ref{tab:kernel pinball comparison quantile-wise}, \ref{tab:kernel mae comparison} and figures \ref{fig:kernel quantile regressors comparison} comparing various kernel functions.

>>>>

We saw in the previous chapter how point forecasting relates to predicting the expected value of the target variable.
Conversely, in probabilistic forecasting we are interested in modelling a probability distribution over the target variable.
This approach enables us to account for prediction uncertainty and to make more informed decisions.
\section{Quantile regression}
\input{input_files/qr.tex}

\section{Quantile forest}
Meinshausen \cite{meinshausen2006quantile} extends the idea of random forest \cite{breiman2001random} generalising it, the result is the quantile forest algorithm. Quantile forest allows us to estimate conditional quantiles in a non parametric fashion.
\\
In order to understand this algorithm, it is first necessary to cover the theory of decision trees and random forests.
\subsection{Decision trees}
Decision trees methods partition recursively the feature space in a set of binary rectangles and then fit a simple model in each of those partitions (the most straightforward is fitting just a constant).
To get started, we first split the space into two disjoint regions, then we model the response variable by the mean of the observed predicted variables with associated features falling in that specific region. Our goal is selecting the best features and best split point such that to achieve the best generalizing fit.
For a visualisation consider the pictures \ref{fig:elements_statistical_learning1} and \ref{fig:elements_statistical_learning2} taken from \cite{hastie2009elements}, where the decision tree algorithm is visualised for a regression problem with two independent variables $X_2$ and $X_1$.
\begin{figure}
    \includegraphics[width=0.5\textwidth]{images/elsii1.png}
    \caption{two-dimensional feature space partitioned by recursive binary splitting \cite{hastie2009elements}}
    \label{fig:elements_statistical_learning1}
  \end{figure}
\\
\begin{figure}
\includegraphics[width=1\textwidth]{images/elsii2.png}
\caption{partition tree and regression model \cite{hastie2009elements}}
\label{fig:elements_statistical_learning2}
\end{figure}
\\
Suppose now to partition the feature space into M regions $R_1,\dots,R_M$, then the model reads as follows.
\begin{equation}
    f(x)=\sum\limits_{m=1}^{M}c_m\mathbb{I}_{\{x \in R_m\}}
\end{equation}
\\
It follows that the function minimising the sum of squares is the one with $\hat{c}_m=mean(y_i|x_i \in R_m)$.
Finding the best split in terms of minimum sum of squares is computationally infeasible in practice. Therefore we approximate a solution by approaching the problem in a greedy fashion.
Let $j$ denote the splitting variable and $s$ be the the split point we define the two half planes
\begin{equation}
    R_1(j,s)=\{X|X_j\leq s\} \quad R_2(j,s)=\{X|X_j >s\}
\end{equation}
Then we search for the $s$ and $j$ that solve
\begin{equation}
    \min_{j,s}\left[\min_{c_1} \sum\limits_{x_i \in R_1(j,s)}(y_i-c_1)^2+\min_{c_2} \sum\limits_{x_i \in R_2(j,s)}(y_i-c_2)^2\right]
\end{equation}
The inner problem is easy, as already pointed out, we will have 
\begin{equation}
    \hat{c}_1=mean(y_i|x_i \in R_1) \quad \hat{c}_2=mean(y_i|x_i \in R_2)
\end{equation}
For the outer problem, we scan through all the (j,s) tuples and pick the best pair. 
Next, one or both of these regions from the previous step are split into two more regions; we recurse this process until some stopping condition is triggered (max number of branches, max deth of tree, threshold on the mean squared error (MSE), minimum number of observations in each leaf node).
Notice, this being a greedy algorithm implies that our final solution is guaranteed to be just a local optimum not a global one.
Even though their simplicity, these models have proved themselves to be really powerful. See figure \ref{fig:decision_tree} for a example where decision tree with different hyperparameters are fitted to a sine wave plus noise. The most popular among these model is the CART \cite{breiman2017classification} tree, its name comes from the fact that it can handle both classification and regression problems.
\begin{figure}
    \includegraphics[width=\textwidth]{images/decision_tree.png}
    \caption{decision tree regression}
    \label{fig:decision_tree}
    % image generated on coolab to be quick
\end{figure}

@@

% \section{Kernel methods}
\section{Kernel quantile regression}
The idea of quantile regression has been extended to kernel methods by Takeuchi et al. \cite{takeuchi2006nonparametric}.
There, they minimize a risk functional plus regularizer defined as follows.
\begin{equation}\label{eq:kqr_min1}
    R[f]:=\frac{1}{m}\sum\limits_{i=1}^{m}\rho_q(y_i-f(x_i))+\frac{\lambda}{2}\|w\|_\mathcal{H}^2
\end{equation}
where $f=w+b$, $w \in \mathcal{H}$ and $b \in \mathbb{R}$.
Using the link between RKHS and feature spaces, we can rewrite $f(x)=\langle w, \phi(x) \rangle_{\mathcal{H}}+b$. 
% Moreover, note that the RKHS norm is defined as follows $\|f\|_{\mathcal{H}}=\inf\{\|w\|_{\mathcal{F}}:w\in \mathcal{F},f(x)=\langle w,\varphi (x)\rangle _{\mathcal{F}},\forall x\in X\}$, where $\mathcal{F}$ is the feature space.
 Doing so we obtain a minimization problem equivalent to minimizing equation \ref{eq:kqr_min1}.
\begin{equation}\label{eq:kqr_min2}
    \begin{aligned}
    \min_{w,b} \quad & C \sum \limits_{i=1}^{m}
    q(y_i-\langle w, \phi(x_i) \rangle_{\mathcal{H}}-b)+ (1-q)(-y_i+\langle w, \phi(x_i) \rangle_{\mathcal{H}}+b)+ \frac{1}{2}\|w\|_{\mathcal{H}}^2\\
    \end{aligned}
    \end{equation}
Note the division by $\lambda$ so that $C:=\frac{1}{\lambda m}$.
\\
We can next rephrase the optimisation in \ref{eq:kqr_min2} by introducing the slack variables $\xi_i$ and $\xi_i^*$.
\begin{equation}\label{eq:kqr_min3}
    \begin{aligned}
        \min_{w,b,\xi_i,\xi_i^*} \quad & C \sum \limits_{i=1}^{m}
        q \xi_i+ (1-q)\xi_i^*+ \frac{1}{2}\|w\|_{\mathcal{H}}^2\\
    \textrm{s.t.} \quad & y_i-\langle w, \phi(x_i) \rangle_{\mathcal{H}}-b \leq \xi_i\\
    & -y_i+\langle w, \phi(x_i) \rangle_{\mathcal{H}}+b \leq \xi_i^*\\
      &\xi_i\geq0    \\
      &\xi_i^*\geq0    \\
    \end{aligned}
    \end{equation}
In order to make it more compact, we rewrite equation \ref{eq:kqr_min3} in matrix notation.
\begin{equation}\label{eq:kqr_min4}
    \begin{aligned}
        \min_{w,b\xi,\xi^*} \quad & C q \xi^\intercal \mathbb{1}+ C q (\xi^*)^\intercal \mathbb{1}+ \frac{1}{2}w^\intercal w\\
    \textrm{s.t.} \quad & y-\Phi^\intercal w -b \preceq \xi\\
    & -y+\Phi^\intercal w +b \preceq \xi^*\\
      &\xi\succeq0    \\
      &\xi^*\succeq0    \\
    \end{aligned}
    \end{equation}
Consider now, the lagrangian $\mathcal{L}$ associated to \ref{eq:kqr_min4}
\begin{equation}\label{eq:kqr_min5}
    \begin{aligned}
    L(w,b,\xi,\xi^*)= & C q \xi^\intercal \mathbb{1}+ C q (\xi^*)^\intercal \mathbb{1}+ \frac{1}{2}w^\intercal w- \lambda^\intercal(\xi - y+\Phi^\intercal w +b)
    - (\lambda^*)^\intercal(\xi^* +y-\Phi^\intercal w -b)
    \\
    & -\nu^\intercal \xi - (\nu^*)^\intercal \xi^*
\end{aligned}
\end{equation}
The next step is deriving its dual formulation, since it is easier and more efficient to solve. This because the dual problem has the useful property of being always convex.

@@

    The dual function associated to the lagrangian $\mathcal{L}(x,\lambda, \nu)$ is given by $g(\lambda, \nu)=\underset{x}\inf \ \mathcal{L}(x,\lambda, \nu)$

@@

where $\lambda$ is called the lagrange multiplier of the optimization problem. Such dual formulation has an useful property, which is \begin{equation}\label{weak_duality}
    g(\lambda, \nu)\leq p^*

@@

where $p^*$ is the optimal value of your optimization problem. In other words $g(\lambda, \nu)$ is a lower bound for the optimal $p^*$.
Consider now a simple lagrangian $\mathcal{L}(x,\lambda, \nu)=f(x)+\sum\limits_{i=1}^n \lambda_i r_i(x) +\sum\limits_{i=1}^{n} \nu_i h_i(x)$, where $r_i(x)$ are inequality constraints while $h_i(x)$ are equality constraints of the problem. Then it can be noted that, the lower bound on $p^*$ is non trivial only when the lagrange multiplier $\lambda \succeq 0$.
% when lagrange multiplier is negative, we have that g > p
Therefore, the idea is that by maximizing the dual function subject to the contraint $\lambda \succeq 0$, we can obtain an approximate or perfect solution to the primal problem.

@@

\begin{proposition}
    Slater's condition reads as

@@

    Where relint $D$ is the relative interior of $ D:=\underset{i=0}{\overset{m}{\cap}} \operatorname {dom} (r_{i})$
\end{proposition}

@@

In our case we don't have any equality constraint, so we do not have to worry about the $h_i(x)=0$ term in \ref{slater_condition}. All we have to check is the convexity of our problem and that there exist an $x$ such that $r_i(x)<0$.

@@

This condition is easily checked by the fact that kernel matrices are by definition positive semidefinite.

@@

Next we check that Slater's condition holds. Considering first the two non negative constraints on $\xi$ and $\xi^*$, we conclude that $\xi$  and $\xi^*$ have to be greater or equal to zero for the existence of an $x$ satisfying Slater's condition. Thus, let us suppose that $0 \leq \xi \leq \mu$ and $0 \leq \xi^* \leq \mu$.

@@

        w=& \Phi^\intercal(\Phi \Phi^\intercal)^{-1} (y-b)

@@

We then have for any $\xi > 0$ and $\xi^* > 0$  that

@@

        -\xi + y -\Phi\Phi^\intercal(\Phi \Phi^\intercal)^{-1}(y-b)-b<&0

@@

        -\xi^* - y +\Phi\Phi^\intercal(\Phi \Phi^\intercal)^{-1}(y-b)+b<&0

@@

We end this section with the derivation of the dual problem; that is the convex problem, which we will solve in order to get the quantiles prediction of our quantile kernel algorithm.

@@

        g(\lambda, \lambda^*, \nu, \nu^*)= & \inf_{\xi, \xi^*, w, b} \mathcal{L}(w,b,\xi,\xi^*,\lambda, \lambda^*, \nu, \nu^*)

@@

        \frac{\partial \mathcal{L}}{\partial w}=0 \implies w=\Phi^\intercal(\lambda-\lambda^*)

@@

        \frac{\partial \mathcal{L}}{\partial b}=0 \implies (\lambda-\lambda^*)^\intercal\mathbb{1}=0

@@

        \frac{\partial \mathcal{L}}{\partial \xi}=0 \implies Cq \mathbb{1}-\lambda- \nu=0

@@

        \frac{\partial \mathcal{L}}{\partial \xi^*}=0 \implies C(1-q)\mathbb{1} -\lambda^* -\nu^*=0

@@

As pointed out previously, the lower bound resulting from the dual formulation is non trivial only when the lagrange multipliers $\lambda$ are $\succeq 0$. Looking at the last two equations of the system \ref{eq:lagrange_derivatives}, this implies the following two constraints $\lambda \in [0, Cq\mathbb{1}]$ and $\lambda^*\in [0, C(1-q)\mathbb{1}]$.

@@

% nu and nu^* go away because they are multipliers accounting for equality constraint, which in the optimum have to be satisfied hence are zero => h_i \times \nu=0

@@

        g(\lambda, \lambda^*)=& \xi^\intercal(Cq\mathbb{1}-\lambda -\nu)+(\xi^*)^\intercal(C(1-q)\mathbb{1}-\lambda^*-\nu^*)-(\lambda-\lambda^*)^\intercal \Phi\Phi^\intercal(\lambda-\lambda^*)

@@

        & +(\lambda-\lambda^*)^\intercal y-(\lambda-\lambda^*)^\intercal b+\frac{1}{2}(\lambda-\lambda^*)^\intercal \Phi\Phi^\intercal(\lambda-\lambda^*)

@@

        g(\lambda, \lambda^*)=& 0+0-\frac{1}{2}(\lambda-\lambda^*)^\intercal \Phi\Phi^\intercal(\lambda-\lambda^*)+(\lambda-\lambda^*)^\intercal y-0

@@

        g(\lambda, \lambda^*)=& -\frac{1}{2}(\lambda-\lambda^*)^\intercal \Phi\Phi^\intercal(\lambda-\lambda^*)+(\lambda-\lambda^*)^\intercal y

@@

Defining $\alpha=(\lambda-\lambda^*)$ and letting $K$ the kernel matrix, we have that the dual optimisation problem reads as follows

@@

        \max_{\alpha} \quad & -\frac{1}{2}\alpha^\intercal K\alpha+\alpha^\intercal y\\

@@

    C(q-1)\mathbb{1}\preceq \alpha \preceq Cq\mathbb{1}\\
    &\alpha^\intercal\mathbb{1}=1

@@

        \min_{\alpha} \quad & +\frac{1}{2}\alpha^\intercal K \alpha-\alpha^\intercal y\\

@@

    C(q-1)\mathbb{1}\preceq \alpha \preceq Cq\mathbb{1}\\
    &\alpha^\intercal\mathbb{1}=1

@@

    f(x)=\sum\limits_{i=1}^{n} \alpha_i k(x_i, x)+b

@@

Since our optimisation problem possesses strong duality and it is differentiable in both the objective and the constraint, we have that it must satisfy the Karush Kuhn Tucker conditions (KKT), see section 5.5.3 \cite{boyd2004convex}.
Thanks to the KKT conditions on the primal optimisation problem we have that $f(x_i)=y_i \ \mathrm{for} \ \alpha_i \not \in \{C(q-1), Cq\}$. 

@@

% optimal lambda times optimal x

@@

    \lambda r_i(x)=&0, \quad i=1,\dots,n

@@

    \nabla \mathcal{L}(x)=&0

@@

        \lambda_i(\xi_i-y_i+r_i)=&0

@@

        \lambda_i^*(\xi_i^*+y_i-r_i)=&0

@@

        \nabla \mathcal{L} =&0

@@

        (Cq-\lambda_i) \xi_i=&0

@@

        (C(1-q)-\lambda_i) \xi_i^*=&0

@@

\begin{equation}
\begin{cases}
    \lambda_i=Cq, \ \lambda_i^*=0 & \implies \lambda_i-\lambda_i^*=Cq, \xi_i\leq 0, \xi^*=0\implies \xi_i-y_i+f_i+b=0 \\
    \lambda_i=0, \ \lambda_i^*=C(1-q) & \implies \lambda_i-\lambda_i^*=C(q-1), \xi_i= 0, \xi^*\leq 0 \implies \xi_i^*+y_i-f_i-b=0 \\
    0\leq \lambda_i< Cq, \ 0\leq \lambda_i^*< C(1-q) & \implies \xi_i=0,  \xi^*_i= 0 \implies -y_i+f_i+b=0, \ y_i-f_i-b=0 \\
\end{cases}
\end{equation}

Therefore in order to retrieve b we simply have to choose an index $i$ such that  $\alpha_i \not \in \{C(q-1), Cq\}$
and let
\begin{equation}
    \begin{aligned}
    b=&y_i-\sum\limits_{i=1}^n \alpha_i k(x_i,x)
    \end{aligned}
\end{equation}

@@

% tables
\begin{table}[!h]
\caption{Pinball loss Melbourne data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & KQR \\
    \midrule
    Pinball loss & 11.278895 & 10.317612 & 10.340842 & \textbf{10.031708} \\
    \bottomrule
    \end{tabular}
\end{table}

\begin{table}[!h]
    \caption{Pinball loss quantile-wise Melbourne data}
    \begin{tabular}{lllll}
    \toprule
    Quantiles & Linear qr & Gbm qr & Quantile forest & KQR \\
    \midrule
    0.100000 & 0.710644 & 0.549232 & 0.562888 & \textbf{0.540235} \\
    0.200000 & 1.155014 & 0.938561 & 0.946712 & \textbf{0.903213} \\
    0.300000 & 1.417805 & 1.212671 & 1.222407 & \textbf{1.173803} \\
    0.400000 & 1.540108 & 1.399925 & 1.409293 & \textbf{1.367943} \\
    0.500000 & 1.574957 & 1.517281 & 1.484589 & \textbf{1.455849} \\
    0.600000 & 1.525114 & 1.498608 & 1.495474 & 1.\textbf{447470} \\
    0.700000 & 1.397918 & 1.372183 & 1.362173 & \textbf{1.331503} \\
    0.800000 & 1.170140 & 1.115077 & 1.123195 & \textbf{1.096606} \\
    0.900000 & 0.787195 & \textbf{0.714075} & 0.734112 & 0.715085 \\
    \bottomrule
    \end{tabular}
\end{table}
As already pointed out, the quantile regression with q=0.5 corresponds to the standard regression problem, hence we can compare the proposed methods also in terms of the mean absolute error.
\begin{table}[!h]
\caption{Mean absolute error Melbourne data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & KQR \\
    \midrule
    MAE & 3.253882 & 3.134805 & 3.095041 & \textbf{3.024336} \\
    \bottomrule
    \end{tabular}
\end{table}  
From these tables, we can see that kernel quantile regression outperformed the simple quantile regressor as well as the more complex models like quantile forest and gradient boosting quantile regression for the Melbourne temperatures dataset. Not only kernel quantile regression was the best in terms of total pinball loss but also the best in terms of each quantile pinball loss and mean absolute error.
Comparison has been carried out on further datasets, yielding similar conclusions as the one of above, to know more see \ref{appendix:quantile_regressor_extensive_comparison}.

We conclude this chapter with reporting the same kind of tables \ref{tab:kernel pinball comparison}, \ref{tab:kernel pinball comparison quantile-wise}, \ref{tab:kernel mae comparison} and figure \ref{fig:kernel quantile regressors comparison} comparing various kernel functions.
